Fitness tracking reveals task-specific associations between memory, mental health, and exercise

### 2022-07-18 paper/old.tex

\title{Fitness tracking reveals task-specific associations between
  memory, mental health, and physical activity}
\author[1, $\star$]{Jeremy R. Manning}
\author[1]{Gina M. Notaro}
\author[1]{Esme Chen}
\author[1]{Paxton C. Fitzpatrick}
\affil[1]{Dartmouth College, Hanover, NH}
\affil[$\star$]{Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\newpage
\begin{abstract}
  Physical activity can benefit both physical and mental well-being.
  Different forms of exercise (e.g., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of physical activity might be similarly
  differentiated.  We focused specifically on how different intensities of
  physical activity might relate to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar physical activity habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  physical activity patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.  Taken together, these
  findings provide foundational work for designing physical activity
  interventions that target specific components of cognitive performance
  and mental health by leveraging low-cost fitness tracking devices.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Physical activity can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17, MorrEtal18, GordEtal17, MorrEtal22, HerrEtal10} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

The physical benefits of exercise can be explained by stress-responses
of the affected body tissues. For example, skeletal muscles that are
taxed during exercise exhibit stress responses~\citep{MortEtal09} that
can in turn affect their growth or atrophy~\citep{SchiEtal13}.  By
comparison, the benefits of physical activity on mental health are less direct.
For example, one hypothesis is that physical activity leads to specific
physiological changes, such as increased aminergic synaptic
transmission and endorphin release, which in turn act on
neurotransmitters in the brain~\citep{PaluSchw00}.  Speculatively, if
different physical activity regimens lead to different neurophysiological
responses, one might be able to map out a spectrum of signalling and
transduction pathways that are impacted by a given type, duration, and
intensity of physical activity in each brain region.  For example, prior work
has shown that physical activity increases acetylcholine levels, starting in
the vicinity of the exercised muscles~\citep{ShoeEtal97}.
Acetylcholine is thought to play an important role in memory
formation~\citep[e.g., by modulating specific synaptic inputs from
entorhinal cortex to the hippocampus, albeit in
rodents;][]{PalaEtal21}.  Given the central role that these medial
temporal lobe structures play in memory, changes in acetylcholine
might lead to specific changes in memory formation and retrieval.

In the present study, we hypothesize that (a) different intensities of physical activity will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistent across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and physical activity.

@@

    We ran an online experiment using the Amazon Mechanical Turk (MTurk)
    platform~\citep{GureEtal16}.  We collected data about each participant's fitness and
    physical activity habits, a variety of self-reported measures concerning their
    mental health, and about their performance on a battery of memory
    tasks.

\subsection*{Experiment}
\subsubsection*{Participants}
We recruited experimental participants by posting our experiment as a
Human Intelligence Task (HIT) on the MTurk platform.
We limited participation to MTurk Workers who had been
assigned a ``master worker'' designation on the platform, given to workers
who score highly across several metrics on a large number of HITs,
according to a proprietary algorithm managed by Amazon.  One criterion
embedded into the algorithm is a requirement that master
workers must maintain a HIT acceptance rate of at least 95\%.  We further
limited our participant pool to participants who self-reported that
they were fluent in English and regularly used a Fitbit fitness
tracker device.  A total of 160 workers accepted our HIT in order to
participate in our experiment.  Of these, we excluded all participants
who failed to log into their Fitbit account (giving us access to their
anonymized fitness tracking data), encountered technical issues (e.g.,
by accessing the HIT using an incompatible browser, device, or
operating system), or who ended their participation prematurely,
before completing the full study.  In all, 113 participants
contributed usable data to the study.

@@

\paragraph{Gender, age, and race.}
Of the 113 participants who contributed usable data, 77 reported their gender as female, 35 as
male, and 1 chose not to report their gender.  Participants ranged in
age from 19 to 68 years old (25\textsuperscript{th} percentile: 28.25
years; 50\textsuperscript{th} percentile: 32 years;
75\textsuperscript{th} percentile: 38 years).  Participants reported
their race as White (90 participants), Black or African American (11
participants), Asian (7 participants), Other (4 participants), and
American Indian or Alaska Native (3 participants).  One participant
opted not to report their race.

\paragraph{Languages.}
All participants reported that they were fluent in either 1 or 2
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 1), and that they were ``familiar'' with between 1 and 11
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 2; 75\textsuperscript{th}
percentile: 3).

@@

\paragraph{Reported water and coffee intake.}
Participants reported the number of 8~oz cups of water and coffee they had
consumed prior to accepting the HIT.  Water consumption ranged from 0 to 6 cups (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 3; 75\textsuperscript{th}
percentile: 4).  Coffee consumption ranged from 0 to 4 cups (25\textsuperscript{th} percentile: 0;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 2).

@@

\paragraph*{Survey questions.}  We collected the following demographic
information from each participant: their birth year, gender, highest
(academic) degree achieved, race, language fluency, and language
familiarity.  We also collected information about participants'
health and wellness, including about their vision, alertness, stress, sleep, coffee
and water consumption, location of their residence, activity typically
required for their job, and physical activity habits.

@@

\paragraph*{Naturalistic recall (Fig.~\ref{fig:tasks}b).}
Participants watched a 2.5-minute video clip entitled ``The Temple of
Knowledge.''  The video comprises an animated story told to StoryCorps
by Ronald Clark, who was interviewed by his daughter, Jamilah Clark.
The narrator (Ronald) discusses growing up living in an apartment over
the Washington Heights branch of the New York Public Library, where his
father worked as a custodian during the 1940s.

After watching the video clip, participants were asked to type out
anything they remembered about what happened in the video.  They typed
their responses into a text box, one sentence at a time.  When the
participant pressed the return key or typed any final punctuation mark
(``.'', ``!'', or ``?'') the text currently entered into the box was
``submitted'' and added to their transcript, and the text box was
cleared to prevent further editing of any already-submitted text.
This was intended to prevent participants from retroactively editing
their previous responses.  Participants were given up to 10~minutes to
enter their responses.  After 4~minutes, participants were given the
option of ending the response period early, e.g., if they felt they
had finished entering all the information they remembered.  Each
participant's transcript was constructed from their submitted
responses by combining the sentences into a single document and
removing extraneous whitespace characters.  Following this 4--10-minute free response period, participants were given a series of 10
multiple choice questions about the conceptual content of the story.
All participants received the same questions, in the same order.
Participants also performed a final delayed memory test, where they
carried out the free response recall task a second time, near the end
of the testing session.  This resulted in a second transcript, for
each participant.

@@

After studying all 10 flashcards, participants were given a multiple
choice memory test where they were shown a series of novel
photographs, each depicting one of the 10 words they had learned.
They were asked to select which (of 4 unique options) Gaelic word went
with the given picture.  The 3 incorrect options were selected at
random (with replacement across trials), and the orders in which the
choices appeared to the participant were also randomized.  Each of the
10 words they had learned was tested exactly once.

@@

\subsubsection*{Comparing recent versus baseline measurements.}
We were interested in separating out potential associations between
\textit{absolute} fitness metrics and \textit{relative} metrics.  To
this end, in addition to assessing potential raw (absolute) fitness
metrics, we also defined a simple measure of recent changes in those
metrics, relative to a baseline:
\[
  \Delta_{R, B} m = \frac{B \sum_{i = 1}^R
  m(i)}{R \sum_{i=R + 1}^{R+B}m(i)},
\]
where $m(i)$ is the value of metric $m$ from $i - 1$ days prior to
testing (e.g., $m(1)$ represents the value of $m$ on the day the
participant accepted the HIT, and $m(10)$ represents the value of $m$
9 days prior to accepting the HIT).  We set $R = 7$ and $B = 30$.  In
other words, to estimate recent changes in any metric $m$, we divided
the average value of $m$ taken over the prior week by the average
value of $m$ taken over the 30 days before that.

@@

\subsubsection*{Reverse correlation analyses}
We sought to characterize potential associations between the \textit{dynamics}
of participants' fitness-related activities leading up to the time
they participated in a memory task and their performance on the given
task.  For each fitness-related variable, we constructed a timeseries
matrix whose rows corresponded to timepoints (sampled once per day)
leading up to the day the participant accepted the HIT for our study,
and whose columns corresponded to different participants.  These
matrices often contained missing entries, since different
participants' Fitbit devices tracked fitness-related activities
differently.  For example, participants whose Fitbit devices lacked
heart rate sensors would have missing entries for any heart
rate-related variables.  Or, if a given participant neglected to wear
their fitness tracker on a particular day, the column corresponding to
that participant would have missing entries for that day.  To create
stable estimates, we smoothed the timeseries of each fitness measure
using a sliding window of 1 week.  In other words, for each fitness
measure, we replaced the ``observed
value'' for each day with the average values of that measure (when
available) over the 7-day interval ending on the given day.

@@

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks (Fig.~\ref{fig:tasks}): a random word list learning ``free
recall'' task; a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or physical activity.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/behavior_overview_immediate}
\caption{\textbf{Immediate memory tests.}  \textbf{a. Free recall.}
  Left: probability of recalling each word first as a function of its
  presentation position.  Middle: probability of transitioning between
successively recalling the word presented at position $i$, followed by
word presented at position $i + \mathrm{Lag}$.  Right: probability of
recalling each word as a function of its presentation position.  See
Figure~\frDetail~for additional details.
\textbf{b. Naturalistic recall.}  Top: 2D embedding of a 2.5-min video
clip; each dot reflects a narrative event (red denotes early events
and blue denotes later events).  Bottom: 2D embedding of the averaged
transcripts of participants' recountings of the narrative (dots: same
format as top panel).  The arrows denote the average trajectory
directions through the corresponding region of text embedding space,
for any participants whose recountings passed through that region.
Blue arrows denote statistically reliable agreement across
participants ($p < 0.05$, corrected).  See Figure~\natDetail~for
additional details.  \textbf{c. Foreign language
  flashcards.} Each bar denotes the average proportion of correctly
recalled Gaelic-English word pairs from early (first 3), late (last
3), or all (i.e., all 10) study positions.  See
Figure~\vocabDetail~for additional details.  \textbf{d. Spatial
  learning.}  Average estimation error in shape locations as a
function of the number of shapes.  See Figure~\spatialDetail~for
additional details.  All panels: error bars and error
ribbons denote bootstrap-estimated 95\% confidence intervals.  Shading
(saturation) denotes results for different subsets of participants
assigned based on their task performance (Figs.~\frDetail,
\natDetail, \vocabDetail, and \spatialDetail~provide information about
which performance metrics and values the shading reflects; in general
more saturated colors denote participants who performed better on the
given task.)  In Panel d, participants are grouped in two ways; in the
left panel, participants are grouped according to the $y$-intercepts of regression lines (estimation error as a
function of the number of shapes); in the right panel, participants
are grouped according to the slopes of the same regression lines.}
\label{fig:immediate_behavior}
\end{figure}

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who engaged in particular intensities of physical activity also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to
participants who exercised in a particular way might also affect their
memory performance on the given task, and/or (b) particular physical activity
behaviors could have a causal impact on memory performance.  We
carried out an exploratory analysis whereby we used a bootstrap-based
approach to identify reliable correlations between different aspects
of memory performance (Fig.~\behBehCorr), different aspects of fitness
(Fig.~\fitFitCorr), different demographic attributes
(Fig.~\demDemCorr), and correlations between memory performance,
fitness information, and demographic attributes (Fig.~\allCorr).
Specifically, we sought to identify correlations that were present in
the same direction (i.e., positive or negative) across different
subsets of participants.  For each test, we report the average correlation (taken
across 10,000 subsets of participants, chosen with replacement) and
an associated two-tailed $p$-value, estimated as
\[
p = 2 \times (1 - q),
\]
where $q$ is the proportion of those
10,000 subsets that exhibited correlations in the same direction  (see
\textit{Exploratory correlation analyses}).  When all 10,000 randomly
chosen subsets of participants exhibited correlations in the same
direction (i.e., all positive correlations or all negative
correlations), we report the $p$-value as $p < 0.0001$.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/combined_correlations_main}
\caption{\textbf{Summaries of correlations between behavioral,
    fitness, and mental health measures.}  The reported values in the
  tables reflect correlations between each pair of measures.  Only statistically reliable correlations
  ($p < 0.05$, corrected) are displayed.  \textbf{a. Correlations
    between behavioral and mental health measures.}  We adjusted each
  task's behavioral measure(s) such that more positive values reflect
  better performance on the given task.  We used participants' mean
  recall accuracies to characterize performance on the free recall and
  foreign language flashcards tasks, and mean precisions to
  characterize performance on the naturalistic recall tasks.  We
  characterized performance on the spatial learning task using the
  (inverted and normalized) intercepts and slopes of linear
  regressions on mean estimation errors as a function of the numbers of
  studied shapes (also see Figs.~\ref{fig:immediate_behavior},
  \ref{fig:delayed_behavior},~\frDetail,~\natDetail,~\vocabDetail,
  and~\spatialDetail).  For each mental health measure, more positive
  values denote greater severity of the given measure.  Typical and
  current stress levels were measured by self report.  Mental health
  information was inferred using each participants' list of
  self-reported medications (see \textit{Methods}).  Positive
  correlations indicate that better performance on a given behavioral
  task is associated with more severe mental health phenotypes.
  \textbf{b. Correlations between fitness and mental health measures.}
  For each fitness measure, more positive values denote higher
  observed scores (i.e., higher resting heart rate, more minutes of
  activity or time spent in each heart rate zone, or greater heart
  rate variability).  The mental health measures in this panel were
  treated as in Panel a.  \textbf{c. Correlations between fitness and
    behavioral measures.}  Each measure reflected in this panel was
  treated as in Panels a and b.}
\label{fig:corr_summaries}
\end{figure}

>>>>

\title{Fitness tracking reveals task-specific associations between
  memory, mental health, and exercise}
\author[1, $\star$]{Jeremy R. Manning}
\author[1]{Gina M. Notaro}
\author[1]{Esme Chen}
\author[1]{Paxton C. Fitzpatrick}
\affil[1]{Dartmouth College, Hanover, NH}
\affil[$\star$]{Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\newpage
\begin{abstract}
  Physical exercise can benefit both physical and mental well-being.
  Different forms of exercise (e.g., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of exercise might be similarly
  differentiated.  We focused specifically on how different forms of
  exercise might relate to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar exercise habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  exercise patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.  Taken together, these
  findings provide foundational work for designing exercise
  interventions that target specific components of cognitive performance
  and mental health by leveraging low-cost fitness tracking devices.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Exercise can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

The physical benefits of exercise can be explained by stress-responses
of the affected body tissues. For example, skeletal muscles that are
taxed during exercise exhibit stress responses~\citep{MortEtal09} that
can in turn affect their growth or atrophy~\citep{SchiEtal13}.  By
comparison, the benefits of exercise on mental health are less direct.
For example, one hypothesis is that exercise leads to specific
physiological changes, such as increased aminergic synaptic
transmission and endorphin release, which in turn act on
neurotransmitters in the brain~\citep{PaluSchw00}.  Speculatively, if
different exercise regimens lead to different neurophysiological
responses, one might be able to map out a spectrum of signalling and
transduction pathways that are impacted by a given type, duration, and
intensity of exercise in each brain region.  For example, prior work
has shown that exercise increases acetylcholine levels, starting in
the vicinity of the exercised muscles~\citep{ShoeEtal97}.
Acetylcholine is thought to play an important role in memory
formation~\citep[e.g., by modulating specific synaptic inputs from
entorhinal cortex to the hippocampus, albeit in
rodents;][]{PalaEtal21}.  Given the central role that these medial
temporal lobe structures play in memory, changes in acetylcholine
might lead to specific changes in memory formation and retrieval.

In the present study, we hypothesize that (a) different exercise
regimens will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistant across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and exercise.

@@

    We ran an online experiment using the Amazon Mechanical Turk (MTurk)
    platform~\citep{GureEtal16}.  We collected data about each participant's fitness and
    exercise habits, a variety of self-reported measures concerning their
    mental health, and about their performance on a battery of memory
    tasks.

\subsection*{Experiment}
\subsubsection*{Participants}
We recruited experimental participants by posting our experiment as a
Human Intelligence Task (HIT) on the MTurk platform.
We limited participation to MTurk Workers who had been
assigned a ``master worker'' designation on the platform, given to workers
who score highly across several metrics on a large number of HITs,
according to a proprietary algorithm managed by Amazon.  One criteria
embedded into the algorithm is a requirement that master
workers must maintain a HIT acceptance rate of at least 95\%.  We further
limited our participant pool to participants who self-reported that
they were fluent in English and regularly used a Fitbit fitness
tracker device.  A total of 160 workers accepted our HIT in order to
participate in our experiment.  Of these, we excluded all participants
who failed to log into their Fitbit account (giving us access to their
anonymized fitness tracking data), encountered technical issues (e.g.,
by accessing the HIT using an incompatible browser, device, or
operating system), or who ended their participation prematurely,
before completing the full study.  In all, 113 participants
contributed usable data to the study.

@@

\paragraph{Gender, age, and race.}
Of the 113 participants who contributed usable data, 77 reported their gender as female, 35 as
male, and 1 chose not to report their gender.  Participants ranged in
age from 19--68 years old (25\textsuperscript{th} percentile: 28.25
years; 50\textsuperscript{th} percentile: 32 years;
75\textsuperscript{th} percentile: 38 years).  Participants reported
their race as White (90 participants), Black or African American (11
participants), Asian (7 participants), Other (4 participants), and
American Indian or Alaska Native (3 participants).  One participant
opted not to report their race.

\paragraph{Languages.}
All participants reported that they were fluent in either 1 and 2
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 1), and that they were ``familiar'' with between 1 and 11
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 2; 75\textsuperscript{th}
percentile: 3).

@@

\paragraph{Reported water and coffee intake.}
Participants reported the number of 8~oz cups of water and coffee they had
consumed prior to accepting the HIT.  Water consumption ranged from
0--6 cups (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 3; 75\textsuperscript{th}
percentile: 4).  Coffee consumption ranged from 0--4 cups (25\textsuperscript{th} percentile: 0;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 2).

@@

\paragraph*{Survey questions.}  We collected the following demographic
information from each participant: their birth year, gender, highest
(academic) degree achieved, race, language fluency, and language
familiarity.  We also collected information about participants'
health and wellness, including about their vision, alertness, stress, sleep, coffee
and water consumption, location of their residence, activity typically
required for their job, and exercise habits.

@@

\paragraph*{Naturalistic recall (Fig.~\ref{fig:tasks}b).}
Participants watched a 2.5~minute video clip entitled ``The Temple of
Knowledge.''  The video comprises an animated story told to StoryCorps
by Ronald Clark, who was interviewed by his daughter, Jamilah Clark.
The narrator (Ronald) discusses growing up living in an apartment over
the Washington Heights branch of the New York Public Library, where his
father worked as a custodian during the 1940s.

After watching the video clip, participants were asked to type out
anything they remembered about what happened in the video.  They typed
their responses into a text box, one sentence at a time.  When the
participant pressed the return key or typed any final punctuation mark
(``.'', ``!'', or ``?'') the text currently entered into the box was
``submitted'' and added to their transcript, and the text box was
cleared to prevent further editing of any already-submitted text.
This was intended to prevent participants from retroactively editing
their previous responses.  Participants were given up to 10~minutes to
enter their responses.  After 4~minutes, participants were given the
option of ending the response period early, e.g., if they felt they
had finished entering all of the information they remembered.  Each
participant's transcript was constructed from their submitted
responses by combining the sentences into a single document and
removing extraneous whitespace characters.  Following this 4--10
minute free response period, participants were given a series of 10
multiple choice questions about the conceptual content of the story.
All participants received the same questions, in the same order.
Participants also performed a final delayed memory test, where they
carried out the free response recall task a second time, near the end
of the testing session.  This resulted in a second transcript, for
each participant.

@@

After studying all 10 flashcards, participants were given a multiple
choice memory test where they were shown a series of novel
photographs, each depicting one of the 10 words they had learned.
They were asked to select which (of 4 unique options) Gaelic word went
with the given picture.  The 3 incorrect options were selected at
random (with replacement across trials), and the order in which the
choices appeared to the participant were also randomized.  Each of the
10 words they had learned were tested exactly once.

@@

\subsubsection*{Comparing recent versus baseline measurements.}
We were interested in separating out potential associations between
\textit{absolute} fitness metrics and \textit{relative} metrics.  To
this end, in addition to assessing potential raw (absolute) fitness
metrics, we also defined a simple measure of recent changes in those
metrics, relative to a baseline:
\[
  \Delta_{R, B} m = \frac{B \sum_{i = 1}^R
  m(i)}{R \sum_{i=R + 1}^{R+B}m(i)},
\]
where $m(i)$ is the value of metric $m$ from $i - 1$ days prior to
testing (e.g., $m(1)$ represents the value of $m$ on the day the
participant accepted the HIT, and $m(10)$ represents the value of $m$
9 days prior to accepting the HIT.  We set $R = 7$ and $B = 30$.  In
other words, to estimate recent changes in any metric $m$, we divided
the average value of $m$ taken over the prior week by the average
value of $m$ taken over the 30 days before that.

@@

\subsubsection*{Reverse correlation analyses}
We sought to characterize potential associations between the \textit{dynamics}
of participants' fitness-related activities leading up to the time
they participated in a memory task and their performance on the given
task.  For each fitness-related variable, we constructed a timeseries
matrix whose rows corresponded to timepoints (sampled once per day)
leading up to the day the participant accepted the HIT for our study,
and whose columns corresponded to different participants.  These
matrices often contained missing entries, since different
participants' Fitbit devices tracked fitness-related activities
differently.  For example, participants whose Fitbit devices lacked
heart rate sensors would have missing entries for any heart
rate-related variables.  Or, if a given participant neglected to wear
their fitness tracker on a particular day, the column corresponding to
that participant would have missing entries for that day.  To create
stable estimates, we smoothed the timeseries of each fitness measure
using a sliding window of 1 week.  In other words, for each fitness
measure, we replaced the ``observed
value'' for each day with the average values of that measure (when
available) over the 7 day interval ending on the given day.

@@

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task (Fig.~\ref{fig:tasks}); a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or exercise.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/behavior_overview_immediate}
\caption{\textbf{Immediate memory tests.}  \textbf{a. Free recall.}
  Left: probability of recalling each word first as a function of its
  presentation position.  Middle: probability of transitioning between
successively recalling the word presented at position $i$, followed by
word presented at position $i + \mathrm{Lag}$.  Right: probability of
recalling each word as a function of its presentation position.  See
Figure~\frDetail~for additional details.
\textbf{b. Naturalistic recall.}  Top: 2D embedding of a 2.5~min video
clip; each dot reflects a narrative event (red denotes early events
and blue denotes later events).  Bottom: 2D embedding of the averaged
transcripts of participants' recountings of the narrative (dots: same
format as top panel).  The arrows denote the average trajectory
directions through the corresponding region of text embedding space,
for any participants whose recountings passed through that region.
Blue arrows denote statistically reliable agreement across
participants ($p < 0.05$, corrected).  See Figure~\natDetail~for
additional details.  \textbf{c. Foreign language
  flashcards.} Each bar denotes the average proportion of correctly
recalled Gaelic-English word pairs from early (first 3), late (last
3), or all (i.e., all 10) study positions.  See
Figure~\vocabDetail~for additional details.  \textbf{d. Spatial
  learning.}  Average estimation error in shape locations as a
function of the number of shapes.  See Figure~\spatialDetail~for
additional details.  All panels: error bars and error
ribbons denote bootstrap-estimated 95\% confidence intervals.  Shading
(saturation) denotes results for different subsets of participants
assigned based on their task performance (Figs.~\frDetail,
\natDetail, \vocabDetail, and \spatialDetail~provide information about
which performance metrics and values the shading reflects; in general
more saturated colors denote participants who performed better on the
given task.)  In Panel d, participants are grouped in two ways; in the
left panel, participants are grouped according to the $y$-intercepts of regression lines (estimation error as a
function of the number of shapes); in the right panel, participants
are grouped according to the slopes of the same regression lines.}
\label{fig:immediate_behavior}
\end{figure}

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who exercised in a
particular way also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to
participants who exercised in a particular way might also affect their
memory performance on the given task, and/or (b) particular exercise
behaviors could have a causal impact on memory performance.  We
carried out an exploratory analysis whereby we used a bootstrap-based
approach to identify reliable correlations between different aspects
of memory performance (Fig.~\behBehCorr), different aspects of fitness
(Fig.~\fitFitCorr), different demographic attributes
(Fig.~\demDemCorr), and correlations between memory performance,
fitness information, and demographic attributes (Fig.~\allCorr).
Specifically, we sought to identify correlations that were present in
the same direction (i.e., positive or negative) across different
subsets of participants.  For each test, we report the average correlation (taken
across 10,000 subsets of participants, chosen with replacement) and
an associated two-tailed $p$-value, estimated as
\[
p = 2 \times (1 - q),
\]
where $q$ is the proportion of those
10,000 subsets that exhibited correlations in the same direction  (see
\textit{Exploratory correlation analyses}).  When all 10,000 randomly
chosen subsets of participants exhibited correlations in the same
direction (i.e., all positive correlations or all negative
correlations), we report the $p$-value as $p < 0.0001$.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/corr_MH}
\caption{\textbf{Memory performance differs according to mental health
    measures.}  The reported values in the table reflect correlations
  between each behavioral measure and mental health measure.
  Only statistically reliable correlations ($p < 0.05$, corrected) are displayed. 
  We used participants' mean recall accuracy to characterize
  performance on the free recall and
  foreign language flashcards tasks, and mean precision to
  characterize performance on the naturalistic recall tasks.  We
  characterized performance on the spatial learning task using the
  (inverted and normalized)
  intercepts and slopes of linear regressions on mean estimation
  errors as a function of the number of studied shapes (also see
  Figs.~\ref{fig:immediate_behavior}, \ref{fig:delayed_behavior},
  \frDetail, \natDetail, \vocabDetail, and \spatialDetail).  Typical
  and current stress levels were measured by self report.  Mental
  health information was inferred using each participants' list of
  self-reported medications.}
\label{fig:mh_corr}
\end{figure}

We also found reliable correlations between participants' fitness and
demographic measures and their behaviors in different tasks
(for a complete report, see Fig.~\allCorr).  For example, recent
low-to-moderate-intensity (``fat burn'') 
cardiovascular activity was positively
correlated with immediate ($r = 0.38, p = 0.03$) and delayed ($r =
0.38, p = 0.029$) recall performance on the naturalistic
memory task.  Recent increases in moderate-intensity (``cardio'')
activity over the prior 7 days (relative to the preceding 30 days) was
also positively correlated with immediate naturalistic recall
performance ($r =
0.48, p = 0.003$) and immediate recall performance on the foreign language
flashcards task ($r = 0.43, p = 0.048$).  Recent high-intensity
(``peak'') activity was positively correlated with performance on the
spatial learning task ($r = 0.34, p < 0.0001$), as were recent increases
in high-intensity activity (prior 7 days versus preceding 30 days; $r
= 0.41, p = 0.01$).  Mental health indicators, such as self-reported
stress levels and medications were also associated with differences in
memory (Figs.~\ref{fig:mh_corr}, \allCorr).  For example, self-reported stress levels at the time of test
were negatively correlated with performance on the delayed memory test
for the foreign language flashcards task ($r = -0.29, p = 0.038$),
whereas participants who were medicated for anxiety and depression
tended to perform slightly (but reliably) \textit{better} on the immediate memory test for the
foreign language flashcards task ($r = 0.11, p < 0.0001$).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/weighted_timecourse_summary}
\caption{\textbf{Dynamics of physical activity varies with memory
    performance and mental health measures.}  \textbf{a. Daily step counts.}  Each
  timecourse is weighted by either performance on immediate recall tests
  (left panel) or on delayed recall tests (right panel).  The black
  (baseline) timecourses display the (unweighted) average across all
  participants.  \textbf{b. Daily duration (in minutes) of low-intensity physical
    activity.}  Timecourses are displayed in the same format and color
scheme as those in Panel A.  Analogous timecourses for additional
fitness-related measures may be found in Figures~\activityTimecourse,
\cardioTimecourse, and \sleepTimecourse.  \textbf{c. Timecourses of physical
  activity, weighted by mental health measures.}  The timecourses in
each panel display the average daily step counts (top panel) or
duration of low-intensity activity (bottom panel).  The colored lines
show average activity dynamics weighted by self-reported stress levels
at the start of the experiment (purple) and self-reported ``typical''
stress levels (pink).  The baseline curves (black) display the
average across all participants (re-plotted in Panel C to illustrate
scale differences across panels).  Timecourses for additional mental
health-related and fitness-related measures may be found in
Figures~\activityTimecourseMH, \cardioTimecourseMH, and \sleepTimecourseMH. Error ribbons in all panels denote
the standard error of the mean.  Horizontal lines below each panel's
timecourses denote intervals over which each weighted measure (color) differs
from the unweighted baseline (via a paired sample two-sided $t$-test
of the weighted mean values for each measure within a 30 day window around each timepoint; horizontal lines denote $p <
0.05$, corrected).}
\label{fig:dynamics}
\end{figure}
